---
title: "Exploring the BRFSS data - first project"
output:
  pdf_document: default
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data


```{r load-data}
load("brfss2013.RData")
```
* * *

## Part 1: Data
(The summary is taken from the http://www.cdc.gov/brfss/annual_data/2013/pdf/Overview_2013.pdf)
The Behavioral Risk Factor Surveillance System (BRFSS) is a collaborative project between all of the states in the United States (US) and participating US territories and the Centers for Disease Control and Prevention (CDC).
BRFSS is an ongoing surveillance system designed to measure behavioral risk factors for the adult population (18 years of age and older) residing in the US. All 50 states and the federal territories collect survey data over usually one to three months.
The BRFSS objective is to collect uniform, state-specific data on preventive health practices and risk behaviors that are linked to chronic diseases, injuries, and preventable infectious diseases that affect the adult population. Factors assessed by the BRFSS in 2013 include tobacco use, HIV/AIDS knowledge and prevention, exercise, immunization, health status, healthy days — health-related quality of life, health care access, inadequate sleep, hypertension awareness, cholesterol awareness, chronic health conditions, alcohol consumption, fruits and vegetables consumption, arthritis burden, and seatbelt use.


The dataset under analysis is called 'brfss2013'. This dataset includes information from collected data and weighting values for adjusting results. It contains 491775 observations of 330 variables.

* * *

## Part 2: Research questions

**Research quesion 1:**   

As a first goal, the aim of the analysis is to have a deeper look to one of the chronic health conditions variable ‘addepev2’ – ‘Ever Told You Had A Depressive Disorder’ from the point of view of family types. I want to figure out whether the type of family the people leave have an influence on individuals treating their depressive disorder. Are they going to the doctor faster if they live in big families or not? And if they do it, are they diagnosed with depression more often?
The selected people are constrained to those where the depressive disorder, private residence, number of adults, men, women and the marital status were all provided. No children are taken into analysis. Note that in my terminology the family and household represent the same entities. The number of children are not taken into account.
 
I defined two type of families:
Nuclear: the number of adult man and/or adult women is 1
Big: the number of adult man and/or adult women is more than 3.
Finally, I have one last category ‘All’ – where the number of adults in the household is taken into account.

To figure out how people treat depressive order, I calculated the rate of  ‘Yes’ and ‘No’ answers of the variable(addepev2) in the above mentioned three categories. These rates give a ‘vague’ overview of the presence of depression in different family types comparing to each other and the society as a whole. 

Variables used in the study:  
addepev2: Ever Told You Had A Depressive Disorder -  is the main variable of this study.   
pvtresd1: Private Residence?    
numadult: Number Of Adults In Household   
nummen: Number Of Adult Men In Household   
numwomen: Number Of Adult Women In Household   
marital: Marital Status   

**Research quesion 2:**

My second research question is: "Are the male veterans more prone to the diseases such as heart attack or coronary diseases or the depression opposed to their counterparts without military carrier?"     

My approach has been to calculate the percentage of the "Yes" answers of the veterans and no veterans represented by variable veteran3 - to the following questions: 

Ever Diagnosed With Heart Attack? - variable(cvdinfr4)      
Ever Diagnosed With Angina Or Coronary Heart Disease (cvdcrhd4)    
Ever Told You Had A Depressive Disorder (variable addepev2)    

Variables used are:     

veteran3: Are You A Veteran   
sex: Respondents Sex     
cvdinfr4: Ever Diagnosed With Heart Attack       
cvdcrhd4: Ever Diagnosed With Angina Or Coronary Heart Disease    
addepev2: Ever Told You Had A Depressive Disorder    

**Research quesion 3:**       

The goal of my third exploratory analysis is to figure out the relation of the education level and average weight in different states.      
The selected people are constrained to those where the state, sex, marital status, education level and the weight were all provided.        

Variables used in the study:

X_state: State Fips Code          
sex: Respondents Sex          
marital: Marital Status          
educa: Education Level           
weight2: Reported Weight In Pounds          

* * *

## Part 3: Exploratory data analysis

**Research quesion 1:**
The main objective here is to study the depressive disorder rate in different family types.

First of all, I did some preliminary analysis to figure out the families where the number of women is more than 3. 

An easy way of tabulating the output is to use the group_by() function by the ‘numwomen’ variable along with the summarise().
In the ‘filter’ condition  I used as.numeric(as.character(numwomen)), because in the filter condition I had to specify a number type rather than factor.

```{r women}
my_brfss_women <- brfss2013 %>%
  select(pvtresd1, numadult, nummen, numwomen, marital, addepev2) %>%
  filter(!is.na(addepev2), pvtresd1 == "Yes", !is.na(numwomen), !is.na(marital)) %>%
  filter(as.numeric(as.character((numwomen))) > 3) %>%
  group_by(numwomen) %>%
  summarise(count = n())

my_brfss_women
```
As you can see from the above listing, there are some interesting insigths. For example, there is a family where are 16 adult women.   

Then, I performed the same summary for the adults men. 

```{r men}
my_brfss_men <- brfss2013 %>%
  select(pvtresd1, numadult, nummen, marital, addepev2) %>%
  filter(!is.na(addepev2), pvtresd1 == "Yes", !is.na(nummen), !is.na(marital)) %>%
  filter(as.numeric(as.character((nummen))) > 3) %>%
  group_by(nummen) %>%
  summarise(count = n())

my_brfss_men
```
The results here show up families with even 40 men.

In the next step, I selected the big families. In my terminology, the big families definition is when the number of adult women or men  are greater then 3.
The first filter selects those observations where the 'addepev2', 'nummen', 'numwomen', 'numadult', 'marital' variables have valid values and the people under analysis live in private residence (pvtresd1 == "Yes").   
The second 'filter' is to assure to have both woman and man in the family.   
The third 'filter' is for the condition of number of women or man to be greater than 3.

```{r big}
my_brfss_big <- brfss2013 %>%
  select(pvtresd1, numadult, nummen, numwomen, marital, addepev2) %>%
  filter(!is.na(addepev2), pvtresd1 == "Yes", !is.na(nummen), !is.na(numwomen), !is.na(numadult), !is.na(marital)) %>%
  filter(as.numeric(as.character(nummen)) > 0, as.numeric(as.character(numwomen)) > 0) %>%
  filter(as.numeric(as.character(nummen)) > 3 | as.numeric(as.character(numwomen)) > 3) %>%
  group_by(addepev2) %>%
  summarise(count = n())

my_brfss_big
```


Then, the nuclear families are counted. In that case, the condition is to have one adult women and/or man in the family.

```{r nuclear}
my_brfss_nuclear <- brfss2013 %>%
  select(pvtresd1, numadult, nummen, numwomen, numadult, marital, addepev2) %>%
  filter(!is.na(addepev2), pvtresd1 == "Yes", !is.na(nummen), !is.na(numwomen), !is.na(numadult), !is.na(marital)) %>%
  filter(as.numeric(as.character(nummen)) > 0, as.numeric(as.character(numwomen)) > 0) %>%
  filter(as.numeric(as.character(nummen)) < 2, as.numeric(as.character(numwomen)) < 2) %>%
  group_by(addepev2) %>%
  summarise(count = n())

my_brfss_nuclear
```

Lastly, the same summary is counted for all the population, regardless of the family type.

```{r all}
my_brfss_all <- brfss2013 %>%
  select(pvtresd1, numadult, nummen, numwomen, numadult, marital, addepev2) %>%
  filter(!is.na(addepev2), pvtresd1 == "Yes", !is.na(nummen), !is.na(numwomen), !is.na(numadult), !is.na(marital)) %>%
  group_by(addepev2) %>%
  summarise(count = n())

my_brfss_all
```

Until now, I have three summary tables with "Yes" and "No" answers to the question "Ever Told You Had A Depressive Disorder?" for all the three population groups. For better plotting, the rate is multiplied by 100.
In the next phase, I calculted the rate of "Yes" and "No" answers for all the three groups.
Also, to prepare plotting, I created a category variable called "type". This variable contains the family category type having three possible values: big, nuclear, all. 

```{r big_summary}

my_big <- my_brfss_big %>%
  summarise(rate = as.integer(my_brfss_big[1,2]/my_brfss_big[2,2]*100))

my_big
```

```{r big_mutate}

my_big <- my_big %>%
  mutate(type = "big")
  
my_big
```

```{r nuclear_summary}

my_nuclear <- my_brfss_nuclear %>%
  summarise(rate = as.integer(my_brfss_nuclear[1,2]/my_brfss_nuclear[2,2]*100))

my_nuclear
```

```{r nuclear_mutate}

my_nuclear <- my_nuclear %>%
  mutate(type = "nuclear")

my_nuclear
```


```{r all_summary}

my_all <- my_brfss_all %>%
  summarise(rate = as.integer(my_brfss_all[1,2]/my_brfss_all[2,2]*100))

my_all
```

```{r all_mutate}

my_all <- my_all %>%
  mutate(type = "all")

my_all
```

Now, I have binding the three summary values in one data frame. I use this data frame for plotting.

```{r putting it all together}
my_data <- bind_rows(my_big,my_nuclear, my_all) 

my_data
```
***Plotting***
```{r plotting question 1}
ggplot(data=my_data, aes(x=type, y=rate ,fill=type))+
  geom_bar(width = 0.3, stat="identity",colour="black")  +
  ggtitle("Ever told depressive disorder rate for family types") +
  xlab("Family type") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())+
  theme(legend.title=element_blank())
  

```

Finally, just to remember my research question was 'Is the type of family influences somehow the problems of having depression? And if the depression arose, are the people willing to consult a doctor? 
As from the my_data summary_table and from the plot could be seen that the rate of "Yes/No" answers to the "Ever Told You Had A Depressive Disorder" (variable "appdev2") is higher a little bit in case of big families.
This could be from several reasons:   

* We have much more nuclear than big families. So the rate maybe is not accurate enough   
* It is possible, that people leaving in bigger families consult the doctor earlier. As they have more family members they consult a specialist with a higher probability   
* It is also possible that living in a big family is more stressful comparing to a nuclear family   

There are many explanations to my question, but I feel that much more analysis is needed to answer it accurately.



**Research quesion 2:**
To start with the technical solution, my second research question is: "Are the male veterans more susceptible to the diseases such as heart attack or coronary diseases or the depression opposed to their counterparts without military carrier?"

I treated separately the three diseses under analysis, namely:    
heart attack (cvdinfr4)     
coronary heart disease (cvdcrhd4)    
depression (addepev2).     


**Disease:: heart attack :: cvdinfr4**        

Firstly, I counted the heart attack cases frequency in case of veterans and no veterans. In my summary table I created a frequency variable called "freq_cvdinfr4".
```{r veteran-diseases: cvdinfr4}
my_brfss_veteran_1 <- brfss2013 %>%
  select(sex, veteran3, cvdinfr4) %>%
  filter(sex == "Male", !is.na(veteran3), !is.na(cvdinfr4)) %>%
  group_by(veteran3, cvdinfr4) %>%
  summarise(freq_cvdinfr4=n())

my_brfss_veteran_1
```

Secondly, calculating the total sum grouped by veterans. The result is stored in total_sum_cvdinfr4 variable.

```{r veteran-diseases: calculating the total for cvdinfr4}
my_brfss_veteran_disease_total <- my_brfss_veteran_1 %>%
  group_by(veteran3) %>%
  summarise(total_sum_cvdinfr4 = sum(freq_cvdinfr4))

my_brfss_veteran_disease_total  
```

Adding total_sum_* column to the summary table in order to calculate the percentage.
```{r veteran-diseases: adding total_sum_cvdinfr4}
my_brfss_veteran_1 <- my_brfss_veteran_1 %>%
  mutate(total_sum_cvdinfr4 = ifelse(veteran3 == "Yes", my_brfss_veteran_disease_total$total_sum_cvdinfr4[1], my_brfss_veteran_disease_total$total_sum_cvdinfr4[2]))

my_brfss_veteran_1  
```

Calculating the percentage and added to the summary table. For simplicity, I truncated percentages to integers.
```{r veteran-diseases: adding percentage}
my_brfss_veteran_1 <- my_brfss_veteran_1 %>%
  mutate(percentage_cvdinfr4 = as.integer(freq_cvdinfr4/total_sum_cvdinfr4*100))

my_brfss_veteran_1
```

The results for heart attack are showed up in the summary table. It can be seen that veterans are more likely to have heart attack than their no veteran counterparts: 13% as opposed to 5%.

**Disease:: coronary heart disease :: cvdcrhd4**     
The same steps are applied like in the previouse disease.


```{r veteran-diseases: cvdcrhd4}
my_brfss_veteran_2 <- brfss2013 %>%
  select(sex, veteran3, cvdcrhd4) %>%
  filter(sex == "Male", !is.na(veteran3), !is.na(cvdcrhd4)) %>%
  group_by(veteran3, cvdcrhd4) %>%
  summarise(freq_cvdcrhd4=n())

my_brfss_veteran_2
```

```{r veteran-diseases: calculating the total for cvdcrhd4}
my_brfss_veteran_disease_total <- my_brfss_veteran_2 %>%
  group_by(veteran3) %>%
  summarise(total_sum_cvdcrhd4 = sum(freq_cvdcrhd4))
 
my_brfss_veteran_disease_total 
```


```{r veteran-diseases: adding total_sum_cvdcrhd4}
my_brfss_veteran_2 <- my_brfss_veteran_2 %>%
  mutate(total_sum_cvdcrhd4 = ifelse(veteran3 == "Yes", my_brfss_veteran_disease_total$total_sum_cvdcrhd4[1], my_brfss_veteran_disease_total$total_sum_cvdcrhd4[2]))
  
my_brfss_veteran_2
```

```{r veteran-diseases: adding percentage_cvdcrhd4}
my_brfss_veteran_2 <- my_brfss_veteran_2 %>%
  mutate(percentage_cvdcrhd4 = as.integer(freq_cvdcrhd4/total_sum_cvdcrhd4*100))

my_brfss_veteran_2
```

The results for heart attack are showed up in the summary table. It can be seen that veterans are more likely to have heart attack than their no veteran counterparts: 12% as opposed to 5%.


**Disease:: Depressive Disorder :: addepev2**     
The same steps are applied like in the previouse disease.

```{r veteran-diseases: addepev2}
my_brfss_veteran_3 <- brfss2013 %>%
  select(sex, veteran3, addepev2) %>%
  filter(sex == "Male", !is.na(veteran3), !is.na(addepev2)) %>%
  group_by(veteran3, addepev2) %>%
  summarise(freq_addepev2=n())

my_brfss_veteran_3
```

```{r veteran-diseases: calculating the total for addepev2}
my_brfss_veteran_disease_total <- my_brfss_veteran_3 %>%
  group_by(veteran3) %>%
  summarise(total_sum_addepev2 = sum(freq_addepev2))
 
my_brfss_veteran_disease_total 
```


```{r veteran-diseases: adding total_sum_addepev2}
my_brfss_veteran_3 <- my_brfss_veteran_3 %>%
  mutate(total_sum_addepev2 = ifelse(veteran3 == "Yes", my_brfss_veteran_disease_total$total_sum_addepev2[1], my_brfss_veteran_disease_total$total_sum_addepev2[2]))

my_brfss_veteran_3  
```

```{r veteran-diseases: adding percentage_addepev2}
my_brfss_veteran_3 <- my_brfss_veteran_3 %>%
  mutate(percentage_addepev2 = as.integer(freq_addepev2/total_sum_addepev2*100))

my_brfss_veteran_3
```
The results for depressive disorder are showed up in the summary table. It can be seen that veterans are slightly susceptible to have depressive disorder than their no veteran counterparts (15% as opposed to 13%).

**Putting it all together :: overall summary table**
```{r veteran-diseases:: putting it all together}
my_brfss_veteran <-cbind(my_brfss_veteran_1, my_brfss_veteran_2, my_brfss_veteran_3)
my_brfss_veteran <- my_brfss_veteran[ -c(6, 11) ]

my_brfss_veteran
```

**Preparing the plot**  
**Putting it all together**   

In order to represent my summary data in a bar plot of three variables representing the three diseases, I created a categorical variable called "disease" having values: cvdinfr4, cvdcrhd4, addepev2.      
Additionally, I created the categorical variable percentage, where are store the percentage values.   

```{r veteran-diseases:: putting it all together for plot}
my_veteran1 <-my_brfss_veteran_1 %>%
  mutate(disease = "cvdinfr4", percentage = percentage_cvdinfr4)
my_veteran2 <-my_brfss_veteran_2 %>%
  mutate(disease = "cvdcrhd4", percentage = percentage_cvdcrhd4)
my_veteran3 <-my_brfss_veteran_3 %>%
  mutate(disease = "addepev2", percentage = percentage_addepev2)
```

From my summary tables I filtered the "Yes" cases of diseases.

```{r veteran-diseases:: plot}
my_veteran1 <-my_veteran1 %>%
  filter(cvdinfr4 == "Yes") %>%
  select(veteran3, disease, percentage) 

my_veteran1
  
my_veteran2 <-my_veteran2 %>%
  filter(cvdcrhd4=="Yes") %>%
  select(veteran3, disease, percentage) 

my_veteran2

my_veteran3 <-my_veteran3 %>%
  filter(addepev2=="Yes") %>%
  select(veteran3, disease, percentage) 

my_veteran3  
```

All my summaries are put binded in an overall summary table.

```{r veteran-diseases: final for plot}

my_data <- rbind(my_veteran1, my_veteran2, my_veteran3)

my_data
```

***Conclusions***

The final results show that male veterans are prone to diseases like heart attack and coronary heart disease much more than their male counterparts without military past. As for depressive disorder there is no such clear evidence as the percentage of the close to each other (15%/13%).   

```{r veteran-diseases: plotting-disease}

ggplot(data=my_data, aes(x=veteran3, y=percentage ,fill=disease))+
  geom_bar(width = 0.3, stat="identity",colour="black")  +
  ggtitle("Diseases for veterans") +
  xlab("Veteran or not?")+ylab("Prone to disease(percent)")
  
```

```{r veteran-diseases: plotting-veteran}

ggplot(data=my_data, aes(x=disease, y=percentage ,fill=veteran3))+
  geom_bar(width = 0.3, stat="identity",colour="black")  +
  ggtitle("Diseases for veterans") +
  xlab("Disease")+ylab("Prone to disease(percent)")
  
```


**Research quesion 3:**         
Just to remember, the purpose of the analysis to figure out if there some relation between obesity and education level.    
To start with, I did a summary count grouped by education level. I selected just the observetions where the relevant varaibles (X_state, sex, marital, educa, weight2) were provided.

```{r - obesity initial view}
my_brfss_obesity <- brfss2013 %>%
  select(X_state, sex, marital, educa, weight2) %>%
  filter(!is.na(X_state), !is.na(sex), !is.na(marital), !is.na(educa), !is.na(weight2)) %>%
group_by(educa) %>%
summarise(count=n()) 

my_brfss_obesity
```
As you can se from the above summary the education level(educa) is a categorical variable having six values.      

In order to be able to have the statistical values for weights I have to transform the weight2 factor variable to numeric variable. This is necessary, as the statistical functions work only on numbers.           
Additionally I change the name of the "educa" variable to "education" and "X_state" to state just to ease my life when plotting the final insights.

```{r - obesity, selecting the required subset}
my_obesity <- brfss2013 %>%
  select(X_state, sex, marital, educa, weight2, height3) %>%
  filter(!is.na(X_state), !is.na(sex), !is.na(marital), !is.na(educa)) %>%
  filter(!is.na(weight2), !is.na(height3))

```

```{r - obesity change variable names}
my_obesity <- my_obesity %>%
  mutate(weight = as.numeric(as.character(weight2)))
my_obesity <- my_obesity %>%
  mutate(state = X_state)
my_obesity <- my_obesity %>%
  mutate(education = educa)
```


```{r - obesity weight statistics}
my_obesity_stat <- my_obesity %>%
   filter(!(is.na(weight))) %>%
  summarise(weightmean = mean(weight), weightmedian = median(weight), weightsd = sd(weight), weightmin = min(weight), weightmax = max(weight))

my_obesity_stat
```


As you can see from the statistics there are some outliers. I filter for 
```{r - outliers}
my_obesity_outliers<-my_obesity %>%
  select(state, education, weight) %>%
   filter(!(is.na(weight))) %>%
  filter(weight>9000) %>%
  group_by(state)
```
We can see that there are 831 observations where weight>9000. I assume that is a mistake. So I omit these observations from futher studies.

I do the statistics one again.

```{r - obesity weight statistics without outliers}
my_obesity_stat <- my_obesity %>%
   filter(!(is.na(weight))) %>%
  filter(weight<1000) %>%
  summarise(weightmean = mean(weight), weightmedian = median(weight), weightsd = sd(weight), weightmin = min(weight), weightmax = max(weight))

my_obesity_stat
```
Now I have more realistic statistics.    

After that, I calculated the mean grouped by state and education.

```{r - obesity education, state}
my_obesity_stedu <- my_obesity %>%
   filter(!(is.na(weight))) %>%
   filter(weight<1000) %>%
   group_by(state, education) %>%
   summarise(weightmean = as.integer(mean(weight)))

my_obesity_stedu
```

Finally, the summary is grouped by education level. 
```{r - obesity education}
my_obesity_edu <- my_obesity %>%
   filter(!(is.na(weight))) %>%
  filter(weight<1000) %>%
   group_by(education) %>%
   summarise(weightmean = as.integer(mean(weight)))

my_obesity_edu
```

***Plotting the result summary***   
```{r}
ggplot(data=my_obesity_edu, aes(x=education, y=weightmean,fill=education))+
  geom_bar(width = 0.7, stat="identity",colour="black")  +
  ggtitle("Weigths per education") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
  

```

As a conclusion, according to these simple statistics it seems that there is no association between education level and obesity. More detailed analysis is needed to have accurate insights.

***Comparing the average weights of two states***
As an example I used Nebraska and Alaska.

```{r}
my_obesity_twostates <- my_obesity %>%
   filter(!(is.na(weight))) %>%
   filter(weight<1000) %>%
   filter(state =="Nebraska"| X_state =="Alaska") %>%
   group_by(education, state) %>%
   summarise(weightmean = as.integer(mean(weight)))

```


```{r}
ggplot(data=my_obesity_twostates, aes(x=education, y=weightmean,fill=state))+
  geom_bar(width = 0.8, stat="identity", position=PositionDodge,colour="black")  +
  ggtitle("Weigths per education level") +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
  

```

Comparing two states there are no semnificative differences.


