---
title: "R Notebook"
output:
  pdf_document: default
  html_notebook: default
---

Let's load the packages

```{r load-packages, message=FALSE}
library(devtools)
library(statsr)
library(dplyr)
library(ggplot2)
```


**6.3 Orange tabbies.** 
Suppose that 90% of orange tabby cats are male. Determine if the following statements are true or false, and explain your reasoning.
(a) The distribution of sample proportions of random samples of size 30 is left skewed.    
(b) Using a sample size that is 4 times as large will reduce the standard error of the sample proportion by one-half.    
(c) The distribution of sample proportions of random samples of size 140 is approximately normal.                       
(d) The distribution of sample proportions of random samples of size 280 is approximately normal.                            

**Solution**


```{r - 6.3}
n_30 <- 30
p <- 0.8
s_cond <- p*n_30
f_cond <- (1-p)*n_30
SE_30 <- sqrt(p*(1-p)/n_30)

n_120 <- 120
SE_120 <- sqrt(p*(1-p)/n_120)
SE_30/SE_120
```


(a) False, as the p = 0.9, sample size is small then the distribution is left skewed. 

Succes - failure condition is not met if n=30. 

(b) True
(c), (d) True

**6.5 Prop 19 in California.**                  
In a 2010 Survey USA poll, 70% of the 119 respondents between the ages of 18 and 34 said they would vote in the 2010 general election for Prop 19, which would change California law to legalize marijuana and allow it to be regulated and taxed. At a 95% confidence level, this sample has an 8% margin of error. Based on this information, determine if the following statements are true or false, and explain your reasoning.    
(a) We are 95% confident that between 62% and 78% of the California voters in this sample support Prop 19.   
(b) We are 95% confident that between 62% and 78% of all California voters between the ages of 18 and 34 support Prop 19.       
(c) If we considered many random samples of 119 California voters between the ages of 18 and 34, and we calculated 95% confidence intervals for each, 95% of them will include the true population proportion of 18-34 year old Californians who support Prop 19.           
(d) In order to decrease the margin of error to 4%, we would need to quadruple (multiply by 4) the sample size.
(e) Based on this confidence interval, there is sufficient evidence to conclude that a majority of California voters between the ages of 18 and 34 support Prop 19.              

**Solution**       
(a) False, as the confidence interval is constructed to estimate the population proportion not the the sample proportion.           

```{r - 6.5}
n <-119
SE <- 0.08
p <- 0.7

CI_l <- 0.7 - 0.08
CI_u <- 0.7 + 0.08
```

(b) True      
(c) True, by the definition of the CI
(d) True, quadrupling the sample size will reduce SE by 
\[ \frac{1}{\sqrt{4}} \]

($\frac{1}{\sqrt{4}}$)     
(e) True. The 95% CI is entirely above 50%.    
      
    
    
**6.9 Life after college.**
We are interested in estimating the proportion of graduates at a mid-sized
university who found a job within one year of completing their undergraduate degree. Suppose we conduct a survey and find out that 348 of the 400 randomly sampled graduates found jobs. The graduating class under consideration included over 4500 students.          
(a) Describe the population parameter of interest. What is the value of the point estimate of this parameter?          
(b) Check if the conditions for constructing a confidence interval based on these data are met.         
(c) Calculate a 95% confidence interval for the proportion of graduates who found a job within one year of completing their undergraduate degree at this university, and interpret it in the context of the data.           
(d) What does \95% confidence" mean?           
(e) Now calculate a 99% confidence interval for the same parameter and interpret it in the context of the data.           
(f) Compare the widths of the 95% and 99% confidence intervals. Which one is wider? Explain.             

$\hat{p}$.

```{r - 6.9}
n <- 400
n_s <- 348
z_score <- 1.96
p_hat <- n_s/n
p_hat

s_verif <- p_hat*n
f_verif <- (1-p_hat)*n

SE <- sqrt(p_hat*(1-p_hat)/n)
  
confidence_interval <- 
  data.frame(lower = p_hat - z_score * SE,
            upper = p_hat + z_score * SE)

confidence_interval
```

(a) Proportion of graduates from this university who found a job within one year of graduating: $\hat{p}=$ `r p_hat`   

(b)   
* Independence verification     
      + Random sample from less than 10% of population.     
* Success-failure verification    
      + Well above 10:  `r s_verif` and `r f_verif`

(c) see above

**6.11 Study abroad**    
A survey on 1,509 high school seniors who took the SAT and who completed an optional web survey between April 25 and April 30, 2007 shows that 55% of high school seniors are fairly certain that they will participate in a study abroad program in college.     
(a) Is this sample a representative sample from the population of all high school seniors in the US? Explain your reasoning.    
(b) Let's suppose the conditions for inference are met. Even if your answer to part (a) indicated that this approach would not be reliable, this analysis may still be interesting to carry out (though not report). Construct a 90% confidence interval for the proportion of high school seniors (of those who took the SAT) who are fairly certain they will participate in a study abroad program in college, and interpret this interval in context.        
(c) What does 90% confidence" mean?     
(d) Based on this interval, would it be appropriate to claim that the majority of high school seniors are fairly certain that they will participate in a study abroad program in college?        

**Solution**
(a) The study represent the students who took the SAT and completed the online survey. So it is not representative.    
(b)

```{r - 6.11}
n <- 1509
p_hat <- 0.55

SE <-sqrt(p_hat*(1-p_hat)/n)
z_score <- qnorm(0.05, lower.tail = FALSE)

z_score

confidence_interval <- 
  data.frame(lower = p_hat - z_score * SE,
             upper = p_hat + z_score * SE)

confidence_interval


```

We are 90% confident that 53% to 57% SAT students are fearly cetain that they will go abroad for study.

(c)    
90% of such random samples would produce a 90% confidence interval that will contain the true proportion.
(d)    
Yes. The interval lies entirely above 50%.

**6.13 Public option, Part I**
A Washington Post article from 2009 reported that support for a government-run health-care plan to compete with private insurers has rebounded from its
summertime lows and wins clear majority support from the public. More specifically, the article says seven in 10 Democrats back the plan, while almost nine in 10 Republicans oppose it. In-dependents divide 52 percent against, 42 percent in favor of the legislation. (6% responded with other). There were 819 Democrats, 566 Republicans and 783 Independents surveyed.
(a) A political pundit on TV claims that a majority of Independents oppose the health care public option plan. Do these data provide strong evidence to support this statement?
(b) Would you expect a confidence interval for the proportion of Independents who oppose the public option plan to include 0.5? Explain.

**Solution**
(a)   
H0 : $p = 0.5$               
H1 : $p > 0.5$    

```{r - 6.13}
n <- 783
p_ind <- 0.52
p <- 0.5

SE <- sqrt(p_ind*(1-p_ind)/n)
z_score <-  (p_ind - p)/SE
p_value <- pnorm(z_score, lower.tail = FALSE)

```

Since the p_value = `r p_value` we fail to reject the H0. The data do not provide strong evidence thet the the majority of Independents oppose the health care public option plan.
(b)   
Yes, since we do not reject H0.  

**6.15 Browsing on the mobile device. ???????** 
A 2012 survey of 2,254 American adults indicates that 17% of cell phone owners do their browsing on their phone rather than a computer or other device.
(a) According to an online article, a report from a mobile research company indicates that 38 percent of Chinese mobile web users only access the internet through their cell phones.
Conduct a hypothesis test to determine if these data provide strong evidence that the proportion of Americans who only use their cell phones to access the internet is different than the Chinese proportion of 38%.
(b) Interpret the p-value in this context.
(c) Calculate a 95% confidence interval for the proportion of Americans who access the internet on their cell phones, and interpret the interval in this context.

(a)
H0: p = 0.38
HA: $p \ne 0.38$
Independence conditions are met. (Random sample <10% of the population) 
Success- failure conditions are met. 

```{r - 6.15}
n <- 2254
p_ind = 0.17
p <- 0.38
SE <- sqrt(p*(1-p)/n)
z_score <-  (p_ind - p)/SE
p_value <- pnorm(z_score, lower.tail = FALSE)
p_value

```

**6.19 College smokers.**
We are interested in estimating the proportion of students at a university who smoke. Out of a random sample of 200 students from this university, 40 students smoke.
(a) Calculate a 95% confidence interval for the proportion of students at this university who smoke, and interpret this interval in context. (Reminder: Check conditions.)
(b) If we wanted the margin of error to be no larger than 2% at a 95% confidence level for the proportion of students who smoke, how big of a sample would we need?

(a) Independence condtiotions are met (n < 10% of the student populations. Success-failure condtioins are met.)

(b)
ME <- z_score*SE<0.2
1.96x1.96xp(1-p)/n<0.2
n > 1.96x1.96xpx(1-p)/0.2

```{r - 6.19}
n <-200
n_smokers <- 40
p <- n_smokers/n

n*(1-p)
n*p

SE <-sqrt(p*(1-p)/n)
z_score <- qnorm(0.025, lower.tail = FALSE)

z_score

confidence_interval <- 
  data.frame(lower = p - z_score * SE,
             upper = p + z_score * SE)

confidence_interval

n_ <- (1.96*1.96)*p*(1-p)/(0.02*0.02)
n_
```

(a) We are 95% confident that between 14.5% and 25.5% of all students in this university are smoking.  

(b) according to the above calculations the sample size should be at lest n_ = 1537. 

**6.27 Public Option, Part III.**
Exercise 6.13 presents the results of a poll evaluating support for the health care public option plan in 2009. 70% of 819 Democrats and 42% of 783 Independents
support the public option.
(a) Calculate a 95% confidence interval for the difference between (pD ???? pI ) and interpret it in this context. We have already checked conditions for you.
(b) True or false: If we had picked a random Democrat and a random Independent at the time of this poll, it is more likely that the Democrat would support the public option than the Independent.

```{r - 6.27}
nD <- 819
pD <- 0.7
nI <- 783
pI <- 0.42

SE <-sqrt(pD*(1-pD)/nD + pI*(1-pI)/nI)
z_score <- qnorm(0.025, lower.tail = FALSE)

z_score

confidence_interval <- 
  data.frame(lower = (pD-pI) - z_score * SE,
             upper = (pD-pI) + z_score * SE)

confidence_interval

```

(a) We are 95% confident that the proportion of Democrats who support the plan is 23% to 33% higher than the proportion of Independents who do.

**6.29 31 Offshore drilling, Part I. and II.**
A 2010 survey asked 827 randomly sampled registered voters in California "Do you support? Or do you oppose? Drilling for oil and natural gas on the Coast of
California? Or do you not know enough to say?" Below is the distribution of responses, separated based on whether or not the respondent graduated from college.
(a) What percent of college graduates and what percent of the non-college graduates in this sample do not know enough to have an opinion on drilling for oil and natural gas on the Coast of California?
(b) Conduct a hypothesis test to determine if the data provide strong evidence that the proportion of college graduates who do not have an opinion on this issue is different than that of non-college graduates.

H0: $p_G = p_N$
HA: $p_G \ne p_N$
Independence conditions are met. (Random sample <10% of the population) 
Success- failure conditions are met.

```{r - 6.29}
nG <- 438
nN <- 389
nNG <- 104
nNN <- 131
pG <- 104/438
pN <- 131/389
point_est <- pG-pN
p_pool <- 286/827

SE <- sqrt(pG*(1-pG)/nG + pN*(1-pN)/nN)

z_score <- point_est/SE
p_value <- 2*(1-pnorm(z_score, lower.tail = FALSE))
```
Note: z_score is negative that's way we substract it from 1. ????????

```{r - 6.31 ---P-value}
nG <- 438
nN <- 389
nNG <- 154
nNN <- 132
pG <- 154/438
pN <- 132/389
point_est <- pG-pN
p_pool <- 286/827

SE <- sqrt(pG*(1-pG)/nG + pN*(1-pN)/nN)

z_score <- point_est/SE
p_value <- 2*pnorm(z_score, lower.tail = FALSE)
```
Note: z-score is positive we don't substract it from 1. ?????????

**6.41 Open source textbook.** 
A professor using an open source introductory statistics book predicts that 60% of the students will purchase a hard copy of the book, 25% will print it out from
the web, and 15% will read it online. At the end of the semester he asks his students to complete a survey where they indicate what format of the book they used. Of the 126 students, 71 said they bought a hard copy of the book, 30 said they printed it out from the web, and 25 said they read it online.
(a) State the hypotheses for testing if the professor's predictions were inaccurate.
(b) How many students did the professor expect to buy the book, print the book, and read the book exclusively online?
(c) This is an appropriate setting for a chi-square test. List the conditions required for a test and verify they are satisfied.
(d) Calculate the chi-squared statistic, the degrees of freedom associated with it, and the p-value.
(e) Based on the p-value calculated in part (d), what is the conclusion of the hypothesis test?
Interpret your conclusion in this context.

```{r - 6.41}
n <-126

EHC <- 0.6*n
EP <- 0.25*n
EO <- 0.15*n

OHC <- 71
OP <- 30
OO <- 25

chi_square <- ((OHC-EHC)^2)/EHC+ ((OP-EP)^2)/EP+((OO-EO)^2/EO) 
df <- 2
p_value <- pchisq(chi_square,df,lower.tail=FALSE) 

```

**6.43 Rock-paper-scissors.**
Rock-paper-scissors is a hand game played by two or more people
where players choose to sign either rock, paper, or scissors with their hands. For your statistics class project, you want to evaluate whether players choose between these three options randomly, or if certain options are favored above others. You ask two friends to play rock-paper-scissors and count the times each option is played. The following table summarizes the data:

Rock Paper Scissors
43 21 35

Use these data to evaluate whether players choose between these three options randomly, or if certain options are favored above others. Make sure to clearly outline each step of your analysis,
and interpret your results in context of the data and the research question.

```{r - 6.43}
OR <- 43
OP <- 21
OS <- 35

ER <- 33
EP <- 33
ES <- 33

chi_square <- ((OR-ER)^2)/ER+ ((OP-EP)^2)/EP+((OS-ES)^2/ES) 
df <- 2
p_value <- pchisq(chi_square,df,lower.tail=FALSE) 

```

**6.45 Quitters.** 
Does being part of a support group affect the ability of people to quit smoking?
A county health department enrolled 300 smokers in a randomized experiment. 150 participants were assigned to a group that used a nicotine patch and met weekly with a support group; the other 150 received the patch and did not meet with a support group. At the end of the study, 40 of the participants in the patch plus support group had quit smoking while only 30 smokers had quit in the other group.
(a) Create a two-way table presenting the results of this study.
(b) Answer each of the following questions under the null hypothesis that being part of a support group does not affect the ability of people to quit smoking, and indicate whether the expected values are higher or lower than the observed values.
i. How many subjects in the "patch + support" group would you expect to quit?
ii. How many subjects in the "patch only" group would you expect to not quit?
                 Quit Not Quit  total
patch+ support   40   110       150
patch            30   120       150
                 70   230       300
                 


```{r - 6.45}
n <- 300
Erow1col1 <- 150*70/300
Erow2col2 <- 150*230/300 

```

**6.47 Offshore drilling Part III.** The table below summarizes a data set we first encountered in Exercise 6.29 that examines the responses of a random sample of college graduates and non-graduates on the topic of oil drilling. Complete a chi-square test for these data to check whether
there is a statistically signifcant difference in responses from college graduates and non-graduates.

               College 
               Yes No  Total
Support        154 132 286
Oppose         180 126 306
Do not know    104 131 235
               438 389 827
```{r - 6.67}  
O11 <- 154
O12 <- 132
O21 <- 180
O22 <- 126
O31 <- 104
O32 <- 131

E11 <- 286*438/827
E12 <- 286*389/827
E21 <- 306*438/827
E22 <- 306*389/827
E31 <- 235*438/827
E32 <- 235*389/827

chi_square <- ((O11-E11)^2)/E11+ ((O12-E12)^2)/E12+((O21-E21)^2/E21) +
  ((O22-E22)^2)/E22+ ((O31-E31)^2)/E31+((O32-E32)^2/E32)
df <- 2
p_value <- pchisq(chi_square,df,lower.tail=FALSE) 

```

